On Sample Size for Block Bootstrap Confidence Intervals to Have Desired Coverage Rates

### 2023-05-31 Manuscript/manuscript.tex
Update

\begin{abstract}
Block bootstrap is widely used in constructing confidence intervals for
parameters estimated from stationary time series. Theoretically, the method
should provide valid confidence intervals as the length of the time series goes
to infinity. In practice, however, it is necessary to know how large of a finite
sample is required for block bootstrap confidence intervals to work well. This
study aims to answer this question in a simple simulation setting where the data
are generated from a first-order autoregressive process. The empirical coverage
rates of several commonly used bootstrap confidence intervals for the mean,
standard deviation, and the lag-1 autocorrelation are compared. In this 
study, we find that a large sample is needed to recover even a simple
parameter like the mean, and that the selection of interval type is very
important, especially in the case of $\phi$, for which the coverage deteriorates
as the sample size increases.
\jy{put most
  notable findings here. Possibly 1) large sample is needed; percentile
  bootstrap is not working well for rho (can we offer some explanation?).}

@@

The necessary sample size for bootstrap standard errors to provide valid
uncertainty measures in practice has not been extensively studied. For
independent data and non-block bootstrap, \citet{hesterberg2015teachers} notes
that while percentile-based CIs from nonparametric bootstrap
are more accurate than $t$-intervals for larger sample sizes, they are
less accurate for smaller sample sizes. \citet{chernick2009revisiting} finds that the the best choice of method for estimating
the parameter of a distribution is dependent on the sample size, the number of
bootstrap resamples, and the confidence level. In the context of structural equation
modeling, \citet{nevitt2001performance} find
that a sample size of 200--1000 is usually sufficient for interval estimation
using standard nonparametric bootstrap. For dependent data, the necessary
sample
size for bootstrap is even less clear. In the setting of linear regression and focused on estimating the proportion of 
variation attributable to a certain variable, \citet{burch2012nonparametric} finds
that three nonparametric bootstrap methods (Standard, Percentile, and Bias-Corrected and
Accelerated) approach the coverage of a pivotal quantity as $n$ gets larger for an
underlying normal distribution. However, they find that for other distributions, 
coverage may deteriorate as $n$ increases. 
When attempting to create interval estimations for the Pearson correlation 
coefficient of bivariate normal data, \citet{puth2015variety} find bootstrap 
methods to be less accurate than the Fisher's transformation method even 
for a sample of size $n = 100$ and correlation coefficient equivalent to $0$.
Few studies have provided a practical guide on what size is needed
for block bootstrap inference for dependent data.
In a linear regression setting
with
dependent data where the
regression errors are generated from a homoscedastic AR(1) process, 
\citet{goncalves2005bootstrap} find that standard error
estimates from moving 
block bootstrap in small samples may be more accurate than
inference from closed-form asymptotic estimates, but moving block bootstrap 
\jy{did they do block or standard bootstrap?}
\mc{moving block bootstrap}
CIs still do not have sufficient coverage of the target parameter even for sample size
$n = 1024$.
\jy{I added more references to the reference folder that studied coverage of
  bootstrap CIs but not block bootstrap CIs.}

The goal of this paper is to provide some practical recommendations on
necessary sample size for block bootstrap with dependent data, similar to what 
was done for basic bootstrap in \citet{hesterberg2015teachers}. We consider a
simple situation of a stationary time series, where the parameters of
interests are the mean, standard deviation, and the first-order
autocorrelation coefficient. We compare five variants of block bootstrap
CIs from the literature \citep{diciccio1996bootstrap,
  rice2006mathematical}: a basic percentile interval, a percentile
interval centered around the estimate, a bias-corrected and
accelerated ($BC_a$) interval, a standard normal interval, and a Student's $t$
interval. Their
empirical coverage rates at different sample sizes and dependence levels are
compared in a simulation study. The results of this study suggest that
appropriate recovery of temporal dependence parameters is heavily dependent on
the type of interval used.

@@

\paragraph{Centered Bootstrap Percentile CI}
The procedure for constructing a Centered Bootstrap Percentile CI based on standard bootstrap is
described in \citet{singh2008bootstrap}.
The $(1 - \alpha)$ Centered Bootstrap Percentile CI is
\[
(2\hat{\theta}_{n} - \hat\theta_{n, 1 - \alpha/2}^{(b)}, \quad
2\hat{\theta}_{n} - \hat\theta_{n, \alpha/2}^{(b)})).
\]
This CI is necessarily centered around the point estimator $\hat\theta_n$ and
is asymmetric, as different critical values are used to compute
the lower and upper bounds.

@@

\paragraph{Proposed Method CI}
Our Proposed Method CI is centered at the point estimator and uses the
variation
in the bootstrap estimates to construct the error bound. This interval requires
the computation of $\bar\theta_n^{(b)}$, the mean of all bootstrap point
estimates, and $\hat{bias} = \bar\theta_n^{(b)} - \hat\theta_n$.
A $(1 - \alpha)$ CI using our Proposed Method is
\[
(\hat\theta_{n, \alpha/2}^{(b)} - \hat{bias}, \quad
\hat\theta_{n, 1 - \alpha/2}^{(b)} - \hat{bias}).
\]
This CI is necessarily centered around .... and
is asymmetric, as different critical values are used to compute
the lower and upper bounds.

@@

Based on Figure~\ref{fig:mu}, 
Student's $t$ CIs appear to have the greatest coverage rates with respect 
to estimating $\mu$. In fact, for $\phi = -0.4$, Student's $t$ CIs over-cover $\mu$
for all sample sizes, whereas other CIs have coverage rates that
hover around the $95\%$ line at this level of $\phi$. For $\phi \geq -0.2$,
$BC_a$ CIs undercover $\mu$ for all $n$ observed. For $\phi = -0.2$, 
Standard, Percentile, CBP, Proposed Method, and $BC$ CIs seem to 
generally cover $\mu$ at the nominal 
rate when $n \geq 400$. Student's $t$ CIs either recover or over-cover $\mu$
for $n \geq 100$. For $\phi = 0$, or when the 
sample is independently normally distributed, $n = 1600$ seems to 
be enough to recover $\mu$ when using Standard, Percentile, CBP, Proposed Method, or $BC$ 
CIs. Student's $t$ CIs cover $\mu$ at the nominal rate for $n \geq 100$ at 
$\phi = 0$.
For $\phi = 0.2$, $n > 3200$ seems to be necessary to recover $\mu$
at the nominal level when using
any of the CIs, although some CIs, such as Student's $t$, are much closer to the
$95\%$ line than others, most notably $BC_a$.
For $\phi = 0.4$, a much larger sample size appears to be necessary. Once again,
Student's $t$ CIs are the closest to $95\%$ coverage, and $BC_a$ CIs are the furthest.

Based on Figure~\ref{fig:sigma}, 
$BC_a$ CIs greatly undercover $\sigma_x$ for every combination of $n$ and 
$\phi$, observed, although
there is no trend of coverage deterioration, indicating that there may eventually
be a sample size large enough for $\sigma_x$ to be covered at the nominal level.
For $\phi = -0.4$, Standard Normal and Student's $t$ CIs recover $\sigma$ when 
$n = 1600$, but undercover $\sigma$ for $n = 3200$, indicating
that a larger $n$ is most likely necessary to consistently cover $\sigma$ at the 
nominal rate.
The other CI types undercover $\sigma$ for all sample sizes at $\phi = 0.4$.
For $\phi  = -0.2$, all CIs except for $BC_a$ are able to cover $\sigma$ at the 
nominal rate for $n \geq 3200$, and Student's $t$ CIs cover $\sigma$ at the 
nominal rate when $n \geq 400$. When $\phi = 0$, $n \geq 1600$ seems to 
be large enough if Standard CIs are used, and $n \geq 400$ is fine is
Student's $t$ CIs are used. $n \geq 3200$ appears necessary 
for other CIs. For $\phi = 0.2$, 
comfortable coverage of $\sigma$ at the nominal level appears to only be true
for Student's $t$ CIs for $n \geq 3200$, although Standard and CBP CIs. For $\phi = 0.4$, a much larger sample size appears to be necessary. Student's $t$ CIs are the closest to $95\%$ coverage, and $BC_a$ CIs are the furthest.

Based on Figure~\ref{fig:phi}, 
Percentile, CBP, $BC$, and $BC_a$ CIs deteriorate clearly for any level of temporal correlation,
so they should not be used. The deterioration becomes more evident
as the absolute value of $\phi$ increases, regardless of whether $\phi$ is 
negative or positive. When $\phi = -0.4$, Standard and Proposed Method CIs 
recover $\phi$ at the nominal rate for $n \geq 400$. At this $\phi$ level,
Student's $t$ CIs either recover or over-cover $\phi$ for $n \geq 100$. When
$\phi = -0.2$, Standard and Proposed Method CIs recover $\phi$ at $n = 800$ and 
$n = 1600$, 
although they undercover $\phi$ at $n = 3600$, indicating that 
perhaps a larger $n$ is needed if these CIs are uses. However,
Student's $t$ CIs cover $\phi$ at the nominal level for $n \geq 400$. When
$\phi = 0$, Standard and Proposed Method CIs recover $\phi$ relatively comfortably 
at $n = 3200$, and 
$n = 800$ seems to be sufficient for Student's $t$ CIs. For $\phi = 0.2$, Standard
and Proposed Method CIs cover $\phi$ at the nominal rate for $n \geq 1600$, whereas 
Student's $t$ CIs can accomplish this at $n \geq 100$. For $\phi = 0.4$, $n \geq 
800$ is large enough for Standard and Proposed Method CIs. Student's $t$ CIs either
recover or over-cover $\phi$ for all $n \geq 100$.

@@

As shown in Figure~\ref{fig:phi}, Percentile CIs appear to suffer from bias; specifically, they seem to heavily underestimate $\phi$. As shown in Figure~\ref{fig:bc}, $BC$ CIs do seem to correct this bias on average, but the width of the 
CIs is very inconsistent. From Figure~\ref{fig:prop}, we can see that 
$prop$ CIs correct the bias using $\hat{bias} = \bar\theta_n^{(b)} - \hat\theta_n$, which seems to usually be negative, indicating a positive correction from
interval construction in Section~\ref{sec:bbci}.

@@

Again, this analysis \eds{what?} \mc{addressed} relies on the assumption that
there is a size $n$ large enough for the method to work: that is, the method's
performance improves as $n$ increases. Out of the five types of intervals used
in this study, this assumption was found to hold true with respect to
estimating $\phi$ only for Standard, Student's $t$, and Proposed Method CIs, whereas 
Percentile, CBP, $BC$, and $BC_a$ intervals exhibited coverage deterioration as $n$
increased. The Percentile CI's coverage deterioration can be attributed to
bias that is not corrected as $n$ increases. Specifically, as $n$ increases, 
the width of the CI decreases, but because the Percentile CI
underestimates $\phi$, the coverage decreases. The $BC$ CI seems to correct the bias, but the width of the CI seems to be inconsistent. The acceleration 
factor of the $BC_a$ CI seems to fail, as the width of the CI
seems to be inconsistent.

When using Student's $t$ centered intervals and $\phi$ is
unknown, the results of this study suggest that an $n > 3200$ may be
necessary for common practice to estimate simpler parameters like $\mu$ 
or $\sigma_x$. If $\phi$ is already known, a lesser $n$ may be adequate
to
estimate these parameters. For lower absolute values of $\phi$, a smaller
$n$ is sufficient when estimating $\mu$ or $\sigma$. For negative
values of $\phi$, coverage of $\mu$ is higher in general. In fact, 
Student's $t$ seems to over-cover $\mu$ for $\phi = 0.4$, 
meaning that if $\phi$ is known to have a very negative value, another CI
should be used.
However, in real world applications, $\phi$ is
more commonly found to be positive, so Student's $t$ should still be used
if $\phi$ is unknown.
Lastly, to estimate $\phi$, an $n$ of around $800$ using the 
Student's $t$ method
 may be sufficient.
Further investigation may be necessary to see if there are other interval
corrections that fix the coverage deterioration problem for $\phi$.

>>>>

\begin{abstract}
Block bootstrap is widely used in constructing confidence intervals for
parameters estimated from stationary time series. Theoretically, the method
should provide valid confidence intervals as the length of the time series goes
to infinity. In practice, however, it is necessary to know how large of a finite
sample is required for block bootstrap confidence intervals to work well. This
study aims to answer this question in a simple simulation setting where the data
are generated from a first-order autoregressive process. The empirical coverage
rates of several commonly used bootstrap confidence intervals for the mean,
standard deviation, and the lag-1 autocorrelation are compared. In this 
study, we find that a large sample is needed to recover even a simple
parameter like the mean, and that the selection of interval type is very
important, especially in the case of the lag-1 autocorrelation, for which the coverage
deteriorates as the sample size increases when using Percentile, 
%Centered Bootstrap Percentile,
Bias-Corrected, or Bias-Corrected and Accelerated Bootstrap intervals.

\jy{put most
  notable findings here. Possibly 1) large sample is needed; percentile
  bootstrap is not working well for rho (can we offer some explanation?).}

@@

The necessary sample size for bootstrap standard errors to provide valid
uncertainty measures in practice has not been extensively studied. For
independent data and non-block bootstrap, \citet{hesterberg2015teachers} notes
that while percentile-based CIs from nonparametric bootstrap
are more accurate than $t$-intervals for larger sample sizes, they are
less accurate for smaller sample sizes. \citet{chernick2009revisiting} finds that the
the best choice of method for estimating
the parameter of a distribution is dependent on the sample size, the number of
bootstrap resamples, and the confidence level. In the context of structural equation
modeling, \citet{nevitt2001performance} find
that a sample size of 200--1000 is usually sufficient for interval estimation
using standard nonparametric bootstrap. For dependent data, the necessary
sample
size for bootstrap is even less clear. In the setting of linear regression with a
focus on estimating the proportion of 
variation attributable to a certain variable, \citet{burch2012nonparametric} finds
that three nonparametric bootstrap methods (Standard, Percentile, and Bias-Corrected
and
Accelerated) approach the coverage of a pivotal quantity as $n$ gets larger for an
underlying normal distribution. However, they find that for other distributions, 
coverage may deteriorate as $n$ increases. 
When attempting to create interval estimations for the Pearson correlation 
coefficient of bivariate normal data, \citet{puth2015variety} find bootstrap 
methods to be less accurate than the Fisher's transformation method even 
for a sample of size $n = 100$ and correlation coefficient equivalent to 0.
Few studies have provided a practical guide on what size is needed
for block bootstrap inference for dependent data.
In a linear regression setting
with
dependent data where the
regression errors are generated from a homoscedastic AR(1) process, 
\citet{goncalves2005bootstrap} find that standard error
estimates from moving 
block bootstrap in small samples may be more accurate than
inference from closed-form asymptotic estimates, but moving block bootstrap 
\jy{did they do block or standard bootstrap?}
\mc{moving block bootstrap}
CIs still do not have sufficient coverage of the target parameter even for sample size
$n = 1024$.
\jy{I added more references to the reference folder that studied coverage of
  bootstrap CIs but not block bootstrap CIs.}

The goal of this paper is to provide some practical recommendations on
necessary sample size for block bootstrap with dependent data, similar to what 
was done for basic bootstrap in \citet{hesterberg2015teachers}. We consider a
simple situation of a stationary time series, where the parameters of
interests are the mean, standard deviation, and the first-order
autocorrelation coefficient. We compare six variants of block bootstrap
CIs from the literature \citep{diciccio1996bootstrap,
  rice2006mathematical}: a Standard Normal CI, and a Student's $t$ CI, a Percentile
Interval CI, 
%a Centered Bootstrap Percentile Interval, 
a Bias-Corrected CI, a Bias-Correct and Accelerated($BC_a$) CI, and a CI using a new
Proposed Method. Their
empirical coverage rates at different sample sizes and dependence levels are
compared in a simulation study. The results of this study suggest that
appropriate recovery of temporal dependence parameters is heavily dependent on
the type of interval used.

@@

%\paragraph{Centered Bootstrap Percentile CI}
%The procedure for constructing a Centered Bootstrap Percentile CI based on
%standard
%bootstrap is
%described in \citet{singh2008bootstrap}.
%The $(1 - \alpha)$ Centered Bootstrap Percentile CI is
%\[
%(2\hat{\theta}_{n} - \hat\theta_{n, 1 - \alpha/2}^{(b)}, \quad
%2\hat{\theta}_{n} - \hat\theta_{n, \alpha/2}^{(b)})).
%\]
%This CI is necessarily centered around the point estimator $\hat\theta_n$ and
%is asymmetric, as different critical values are used to compute
%the lower and upper bounds.

@@

\paragraph{Proposed Method CI}
Our Proposed Method CI is centered at the point estimator and uses the
variation
in the bootstrap estimates to construct the error bound. This interval requires
the computation of $\bar\theta_n^{(b)}$, the mean of all bootstrap point
estimates, and $\widehat{\text{bias}} = \bar\theta_n^{(b)} - \hat\theta_n$.
A $(1 - \alpha)$ CI using our Proposed Method is
\[
(\hat\theta_{n, \alpha/2}^{(b)} - \widehat{\text{bias}}, \quad
\hat\theta_{n, 1 - \alpha/2}^{(b)} - \widehat{\text{bias}}).
\]
This CI is necessarily centered around .... and
is asymmetric, as different critical values are used to compute
the lower and upper bounds.

@@

Based on Figure~\ref{fig:mu}, 
Student's $t$ CIs appear to have the greatest coverage rates with respect 
to estimating $\mu$. In fact, for $\phi = -0.4$, Student's $t$ CIs over-cover $\mu$
for all sample sizes, whereas other CIs have coverage rates that
hover around the $95\%$ line at this level of $\phi$. For $\phi \geq -0.2$,
$BC_a$ CIs undercover $\mu$ for all $n$ observed. For $\phi = -0.2$, 
Standard, Percentile, %CBP, 
Proposed Method, and $BC$ CIs seem to 
generally cover $\mu$ at the nominal 
rate when $n \geq 400$. Student's $t$ CIs either recover or over-cover $\mu$
for $n \geq 100$. For $\phi = 0$, or when the 
sample is independently normally distributed, $n = 1600$ seems to 
be enough to recover $\mu$ when using Standard, Percentile, 
%CBP, 
Proposed Method, or $BC$ 
CIs. Student's $t$ CIs cover $\mu$ at the nominal rate for $n \geq 100$ at 
$\phi = 0$.
For $\phi = 0.2$, $n > 3200$ seems to be necessary to recover $\mu$
at the nominal level when using
any of the CIs, although some CIs, such as Student's $t$, are much closer to the
$95\%$ line than others, most notably $BC_a$.
For $\phi = 0.4$, a much larger sample size appears to be necessary. Once again,
Student's $t$ CIs are the closest to $95\%$ coverage, and $BC_a$ CIs are the furthest.

Based on Figure~\ref{fig:sigma}, 
$BC_a$ CIs greatly undercover $\sigma_x$ for every combination of $n$ and 
$\phi$, observed, although
there is no trend of coverage deterioration, indicating that there may eventually
be a sample size large enough for $\sigma_x$ to be covered at the nominal level.
For $\phi = -0.4$, Standard Normal and Student's $t$ CIs recover $\sigma$ when 
$n = 1600$, but undercover $\sigma$ for $n = 3200$, indicating
that a larger $n$ is most likely necessary to consistently cover $\sigma$ at the 
nominal rate.
The other CI types undercover $\sigma$ for all sample sizes at $\phi = 0.4$.
For $\phi  = -0.2$, all CIs except for $BC_a$ are able to cover $\sigma$ at the 
nominal rate for $n \geq 3200$, and Student's $t$ CIs cover $\sigma$ at the 
nominal rate when $n \geq 400$. When $\phi = 0$, $n \geq 1600$ seems to 
be large enough if Standard CIs are used, and $n \geq 400$ is fine is
Student's $t$ CIs are used. $n \geq 3200$ appears necessary 
for other CIs. For $\phi = 0.2$, 
comfortable coverage of $\sigma$ at the nominal level appears to only be true
for Student's $t$ CIs for $n \geq 3200$, although Standard %and CBP% 
CIs do reach the nominal level in our simulation results. For $\phi = 0.4$, a much
larger sample size appears to be necessary. Student's $t$ CIs are the closest to
$95\%$ coverage, and $BC_a$ CIs are the furthest.

Based on Figure~\ref{fig:phi}, 
Percentile, %CBP, %
$BC$, and $BC_a$ CIs deteriorate clearly for any level of temporal correlation,
so they should not be used. The deterioration becomes more evident
as the absolute value of $\phi$ increases, regardless of whether $\phi$ is 
negative or positive. When $\phi = -0.4$, Standard and Proposed Method CIs 
recover $\phi$ at the nominal rate for $n \geq 400$. At this $\phi$ level,
Student's $t$ CIs either recover or over-cover $\phi$ for $n \geq 100$. When
$\phi = -0.2$, Standard and Proposed Method CIs recover $\phi$ at $n = 800$ and 
$n = 1600$, 
although they undercover $\phi$ at $n = 3200$, indicating that 
perhaps a larger $n$ is needed if these CIs are uses. However,
Student's $t$ CIs cover $\phi$ at the nominal level for $n \geq 400$. When
$\phi = 0$, Standard and Proposed Method CIs recover $\phi$ relatively comfortably 
at $n = 3200$, and 
$n = 800$ seems to be sufficient for Student's $t$ CIs. For $\phi = 0.2$, Standard
and Proposed Method CIs cover $\phi$ at the nominal rate for $n \geq 1600$, whereas 
Student's $t$ CIs can accomplish this at $n \geq 100$. For $\phi = 0.4$, $n \geq 
800$ is large enough for Standard and Proposed Method CIs. Student's $t$ CIs either
recover or over-cover $\phi$ for all $n \geq 100$.

@@

As shown in Figure~\ref{fig:phi}, Percentile CIs appear to suffer from bias;
specifically, they seem to heavily underestimate $\phi$. As shown in
Figure~\ref{fig:bc}, $BC$ CIs do seem to correct this bias on average, but the width of
  the 
CIs is very inconsistent. From Figure~\ref{fig:prop}, we can see that 
  $prop$ CIs correct the bias using $\widehat{\text{bias}} = \bar\theta_n^{(b)} -
  \hat\theta_n$, which seems to usually be negative, indicating a positive correction
  from
interval construction in \ref{sec:bbci}.

@@

Again, this analysis \eds{what?} \mc{addressed} relies on the assumption that
there is a size $n$ large enough for the method to work: that is, the method's
performance improves as $n$ increases. Out of the five types of intervals used
in this study, this assumption was found to hold true with respect to
estimating $\phi$ only for Standard, Student's $t$, and Proposed Method CIs, whereas 
Percentile, 
%CBP, 
$BC$, and $BC_a$ intervals exhibited coverage deterioration as $n$
increased. The Percentile CI's coverage deterioration can be attributed to
bias that is not corrected as $n$ increases. Specifically, as $n$ increases, 
the width of the CI decreases, but because the Percentile CI
underestimates $\phi$, the coverage decreases. The $BC$ CI seems to correct the bias,
but the width of the CI seems to be inconsistent. The acceleration 
factor of the $BC_a$ CI seems to fail, as the width of the CI
seems to be inconsistent.

When using Student's $t$ intervals and $\phi$ is
unknown, the results of this study suggest that an $n > 3200$ may be
necessary for common practice to estimate simpler parameters like $\mu$ 
or $\sigma_x$. If $\phi$ is already known, a lesser $n$ may be adequate
to
estimate these parameters. For lower absolute values of $\phi$, a smaller
$n$ is sufficient when estimating $\mu$ or $\sigma$. For negative
values of $\phi$, coverage of $\mu$ is higher in general. In fact, 
Student's $t$ seems to over-cover $\mu$ for $\phi = 0.4$, 
meaning that if $\phi$ is known to have a very negative value, another CI
should be used.
However, in real world applications, $\phi$ is
more commonly found to be positive, so Student's $t$ should still be used
if $\phi$ is unknown.
Lastly, to estimate $\phi$, an $n$ of around $800$ using the 
Student's $t$ method
 may be sufficient.
Further investigation may be necessary to see if there are other interval
corrections that fix the coverage deterioration problem for $\phi$.
